A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-12 paper/paper.tex
missing period
edits to related work

\item In \cref{sec:examples} we demonstrate how to use our syntax for
  formal category theory

@@

\section{Discussion}

\paragraph{Calculi for formal category theory}

In their book on $\infty$-category theory, Riehl and Verity use the
formal language of virtual equipments to prove results valid for
$\infty$-categories without concrete manipulation of model
categories. They formalize this language using Makkai's FOLDS
framework of first-order logic with dependent sorts. The approach is
similar in spirit to our \ohol{}, and has the same models. The main
difference is that their theory is built on top of first-order logic,
whereas ours is built on a custom-made variant of higher-order logic
over a (unary) simple type theory. For this reason, statements in
their theory are quite complex as they have to be encoded into
first-order logic, whereas we argue our type-theoretic approach
formalizes informal arguments much more directly, as shown in
Section~\ref{sec:examples}.
%
An informal syntax for manipulating profunctors involving ends and
coends, called the coend calculus is closely related, an extension of
\ohol{} to treat profunctors of many variables of different variances
should provide a formal treatment for the coend calculus
\citet{coend-calculus}.

Another approach to formal category theory is to develop a
\emph{directed dependent type theory}, a type theory where the
identity type is interpreted as morphisms in a (possibly
$\infty$-)category. Research on directed dependent type theories is
ongoing \cite{riehl-shulman,north}. Directed dependent type theories
should generalize \ohol{} in the same way that dependent type theory
generalizes higher order logic: internally there should be a type that
models the category of sets, and terms of this type should model the
profunctors of \ohol{}.  As a foundation for a domain-specific proof
assistant for category theory, \ohol{} may provide a foundation for a
proof assistant in the style of Isabelle/HOL \cite{isabelle-hol},
whereas directed type theory would provide a foundation for a proof
assistant in the style of Agda or Coq \cite{agda,coq}.

@@

One alternative to proarrow equipments is to axiomatize the bicategory
of categories, profunctors and natural transformations rather than the
virtual double category of categories, functors, profunctors and
natural transformations. An abstract structure along
these lines is a \emph{Cartesian bicategory}\footnote{which is not the
same as a bicategory with products/finite limits, for which a syntax
was developed \citet{fioresaville}} is given in
\citet{carboni-walters,carboni-walters-someonelese}. 
Some preliminary work on a proof system for Cartesian bicateogires was
presented at \citet{jonasfreycartesianbicategories}. This bears strong
similarities to our calculus VETT, except that (1) their profunctors
may have 0, 1 or more covariant or contravariant variables but (2)
they do not have a term syntax like our natural transformation syntax
and (3) they do not have functors and restriction of profunctors by
functors (since this structure is not present in a Cartesian
bicategory). Semantically, Cartesian bicategories are somewhat less
general than virtual equipments because in general the set of
functors between two categories can only be recovered from the set of
profunctors if the codomain category is Cauchy complete, in which case
functors are equivalent to adjoint pairs of profunctors.

Finally, some variations on double categories have been used to model
the structure of certain program logics. \citet{newlicata20} present a
logic for \emph{vertically thin} pro-arrow equipments, i.e., pro-arrow
equipments where the functor and natural transformation judgments are
posetal. This calculus is quite different from ours in that the
horizontal morphisms are terms of a programming language, and so they
are not presented as having co-variant and contravariant variables,
and the equipment structure is modeled by explicit coercions rather
than by syntactic substitution of vertical arrows in horizontal
arrows. Another similar calculus is System P \citep{systemP} which is
an internal language of \emph{reflexive graph categories}, which are
like double categories without horizontal composition.

>>>>

\item In \cref{sec:examples} we demonstrate how to use our syntax for
  formal category theory.

@@

\section{Related and Future Work}

We now comapre \vett{} with other calculi for formal category theory.

Riehl and Verity~\cite{rvXXelements} use a formal language of virtual
equipments to prove results valid for $\infty$-categories without
concrete manipulation of model categories. They formalize this language
as a theory in Makkai's framework of first-order logic with dependent
sorts (FOLDS).  While this previous work has the same models as \vett{},
we believe that the syntax we propose in this paper formalizes informal
arguments more directly, as shown in Section~\ref{sec:examples}.  This
is because FOLDS approach approach is entirely relational, whereas we
formalize concepts like restriction of a profunctor or composition of
natural transformations as functional operations (substitution).  In
particular, this means that our calculus requires only vertically
degenerate squares (elements/natural transformations) as a
``user-facing'' notion, with general squares occuring only in the
admissible substitution operations.

%% The main difference is that their theory is built on top of
%% first-order logic, whereas ours is built on a custom-made variant of
%% higher-order logic over a (unary) simple type theory. For this reason,
%% statements in their theory are quite complex as they have to be encoded
%% into first-order logic, whereas we argue our type-theoretic approach
%% %

The coend calculus~\citet{coend-calculus} is an informal syntax for
manipulating profunctors involving ends and coends; an extension of
\vett{} to treat profunctors of many variables of different variances
should would provide a formal treatment of it.

@@

One alternative to proarrow equipments is to axiomatize the bicategory
of categories, profunctors and natural transformations rather than the
virtual double category of categories, functors, profunctors and natural
transformations. An abstract structure along these lines is a
\emph{Cartesian
  bicategory}~\citet{carboni-walters,carboni-walters-someonelese}.  (In
a terminological collision, \citet{fioresaville} give a syntax for a a
bicategory with products/finite limits, but this is a different notion.)
\citet{jonasfreycartesianbicategories} describe some preliminary work on
a proof system for Cartesian bicateogires.  This calculus is more
general than \vett{} in that their profunctors may have 0, 1 or more
covariant or contravariant variables, but they do not have a term syntax
for the natural transformations/elements, and Cartesian bicategories do
not have functors and restriction of profunctors by
functors. Semantically, Cartesian bicategories are somewhat less general
than virtual equipments because in general the set of functors between
two categories can only be recovered from the set of profunctors if the
codomain category is Cauchy complete, in which case functors are
equivalent to adjoint pairs of profunctors.

Our work in this paper fits broadly into a line of work on
\emph{directed dependent type theories}, a type theory where the
identity type is interpreted as morphisms in a (possibly
$\infty$-)category.  In directed type theories based on a bisimplicial
model~\cite{riehl-shulman,weinberger,otherweinbegerpaper,licataweaver},
morphism types are defined using an interval object, like in cubical
type theory~\cite{bch,cchm,afh,abcfhl}, and universal properties like
``morphism induction'' are an internally definable property of certain
types.  Other type theories~\cite{north,ahrensnorth} define morphism
types via an induction principle, corresponding to the lifting
properties of certain kinds of fibrations of categories.  While these
previous works can express some constructions on $\Cat$ that are not
expressible in \vett{}, because \vett{} is more restricted, it has more
models, e.g. in internal and enriched categories, so the theorems that
are provable in \vett{} apply in more settings.

Our use of dependent type theory as a metalanguage, with new judgements
for the virtual equipment structure dependent on the meta-language
context, is similar to the indexed type theories developed by
\citet{isaev}.

%% Directed dependent type theories
%% should generalize \ohol{} in the same way that dependent type theory
%% generalizes higher order logic: internally there should be a type that
%% models the category of sets, and terms of this type should model the
%% profunctors of \ohol{}.  As a foundation for a domain-specific proof
%% assistant for category theory, \ohol{} may provide a foundation for a
%% proof assistant in the style of Isabelle/HOL \cite{isabelle-hol},
%% whereas directed type theory would provide a foundation for a proof
%% assistant in the style of Agda or Coq \cite{agda,coq}.

Finally, some variations on double categories have been used to model
the structure of certain program logics. \citet{newlicata20} present a
logic for \emph{vertically thin} pro-arrow equipments, i.e., pro-arrow
equipments where the functor and natural transformation judgments are
posetal. This calculus is quite different from ours in that the
horizontal morphisms are terms of a programming language, and so they
are not presented as relations with co-variant and contravariant
variables, and the equipment structure is modeled by explicit coercions
rather than by syntactic substitution of vertical arrows in horizontal
arrows. Another similar calculus is System P \citep{systemP} which is an
internal language of \emph{reflexive graph categories}, which are like
double categories without horizontal composition.

%% DRL note: I would compress the rest of this to a couple of sentences
%% as the first thing to remove for space.
